---
title: "Reproducable Research Course Project 1"
author: "Mike G."
date: "4/16/2018"
output: html_document
---

================================================================================
```{r, include=FALSE}
library(knitr)
setwd('~/Documents/HopkinsDataScience/RepData_PeerAssessment1/')
opts_chunk$set(fig.path='./figure/')
``` 

Step 1 is to load the data. This is done with read.table in R. The data (activity)
consists of 3 variables: steps, data, and interval.

```{r , echo=TRUE}
activity = read.table('~/Documents/HopkinsDataScience/RepData_PeerAssessment1/activity.csv', sep=',', header=TRUE, stringsAsFactors = FALSE)

#convert date to a date format
activity$date = as.Date(activity$date, format='%Y-%m-%d')
```

The second step asks for the what are the mean total of steps take per day. To 
do this, we ignore any observations where the number of steps is NA with na.rm.
We pipe the data into group_by and then to mutate to create a totalsteps column.
A histogram is created using ggplot and then the mean and median of totalsteps
are calculated and displayed.

```{r, echo=TRUE}
library(dplyr)
library(ggplot2)
#Calculate total number of steps take per day
#do this with dplyr group_by and summarize and pipe to ggplot to make histogram
activity = activity %>% group_by(date) %>% mutate(totalsteps = sum(steps, na.rm=TRUE))

```{r histogram1}
ggplot(activity, aes(x=totalsteps)) + geom_histogram(bins=20) + 
  xlab('Total steps per day') + ylab('Frequency')
```

```{r}
#calculate mean and median of total steps per day and reprot them
meansteps_na = round(mean(activity$totalsteps, na.rm=TRUE),3)
mediansteps_na = round(median(activity$totalsteps, na.rm=TRUE),3)
```
The mean total steps is `r meansteps_na`.
The median total steps is `r mediansteps_na`.

Step 3: make a time series plot of the 5-minute interval and the average number
of steps taken, averaged across all days

```{r}
#create average number of steps taken across all days by 5-minute interval?
activity_byinterval = activity %>% group_by(interval) %>% mutate(meansteps
                                                      = mean(steps, na.rm=TRUE))
```
```{r stepslinegraph}
activity_byinterval %>% ggplot(aes(x=interval, y=meansteps)) + geom_line() +
  xlab('5-minute interval') + ylab('Mean number of steps')
```
```{r}
rows = which(activity_byinterval$meansteps== max(activity_byinterval$meansteps))
maxinterval = activity_byinterval[rows[1],3]
print(paste('5-minute interval that contains maximum number of steps: ', maxinterval))
```

Step 4: Imputing missing values. The first portion of this is to determine the 
number of missing values in the dataset. Next, I use the median value of the 
interval (e.g., fill in the missing value for interval 5 with the median of the 
steps taken in interval 5 without missing values). Then, a new dataset is 
created with the missing values filled in that has the same number of rows as 
the original dataset. Next, a histogram of the total number of steps taken each 
day. Finally, the mean and median total number of steps taken per day is reported

```{r, echo=TRUE}
#calculate total number of NAs
totalNAs = sum(is.na(activity$steps))
print(paste('Total number missing values: ', totalNAs))

#strategy for filling in missing data (# steps). Will use the median 
activ_med_missing = activity %>% group_by(interval) %>% mutate(mediansteps = median(steps, na.rm=TRUE))

#replace the NAs in steps with the median value computed above
activ_med_missing$steps[is.na(activ_med_missing$steps)] = 
    activ_med_missing$mediansteps[is.na(activ_med_missing$steps)]

activity_nona = activ_med_missing[,-c(4,5)]
activity_nona = activity_nona %>% group_by(date) %>% mutate(totalsteps = sum(steps, na.rm=TRUE))
```

```{r nona_histogram}
#make histogram for new no na dataset
activity_nona %>% group_by(date) %>% ggplot(aes(x=totalsteps)) + geom_histogram()

#report mean and median total steps per day
mean_nona = round(mean(activity_nona$totalsteps),3)
median_nona = round(median(activity_nona$totalsteps),2)

mean_nona
median_nona
```

This shows that the mean number of steps taken per day is greater than that of 
the data with the missing values (`r meansteps_na`, `r mean_nona`). However, the median is the same (`r mediansteps_na`, `r median_nona`).

The final step in this project is to see if there are differences in activity
patterns between weekdays and weekends. First, create a variable 'weekday' that
is the weekday of date in the dataset. Then, convert this to a factor where 1
is a weekday and 0 is a weekend. Next, a dataset is created with no missing values
and by interval using group_by(interval, weekday) and mutate to calculate the
mean steps per interval/weekday. Finally, a plot is created that shows the mean
number of steps by interval/weekday and faceted by weekday.

```{r, echo=TRUE}
#differences in activity patterns between weekdays/weekends?
activity_nona$weekday = weekdays(activity_nona$date)

#create factor variable that distinguishes weekday and weekend (1 = weekday, 0 = weekend)
activity_nona$weekday = gsub('Sunday', 'Weekend', activity_nona$weekday)
activity_nona$weekday = gsub('Saturday', 'Weekend', activity_nona$weekday)
activity_nona$weekday = gsub('Monday', 'Weekday', activity_nona$weekday)
activity_nona$weekday = gsub('Tuesday', 'Weekday', activity_nona$weekday)
activity_nona$weekday = gsub('Wednesday', 'Weekday', activity_nona$weekday)
activity_nona$weekday = gsub('Thursday', 'Weekday', activity_nona$weekday)
activity_nona$weekday = gsub('Friday', 'Weekday', activity_nona$weekday)

activity_nona$weekday = as.factor(activity_nona$weekday)

#panel plot of 5-minute interval and average number of steps across weekdays or weekends
activity_nona_byinterval = activity_nona %>% group_by(interval, weekday) %>% 
  mutate(meansteps = mean(steps))
```

```{r weekday_weekend}
activity_nona_byinterval %>% ggplot(aes(x=interval, y=meansteps)) + geom_line() +
  facet_wrap( ~ weekday, nrow=2, ncol=1) + ylab('Mean number of steps') + xlab('5-minute interval')
```

There is a definite differnce in activity between weekdays and weekends. For 
weekdays, there is a larger spike early on than in weekends. Weeknds show a much
more consistent activity level throughout the day than in weekdays. This is 
probably due to jobs. Weekends tend to be more active during normal working hours.